---
title: "Machine Learning"
author: "John Hamre III"
date: "August 11, 2017"
output: slidy_presentation
---

## Overview

In this presentation I show some of the ways computers can "learn" or be trained, and then predict classifications based off of training sets.

## Bullet Points

- To begin with machine learning I start with knn prediction. It is a powerful and straight forward algorithm to use in R
- PCA clustering is demonstrated next, and is more complicated because it uses eigen values to project multi-dimensional data into two dimensions
- Next, we go into tree-based classification methods of machine learning by looking at rpart and random forest algorithms 
- After an important gene or factor is found it can be used as a measure of prediction
-Finally, I end with clustering as a means of prediction by using k-means and PAM clustering

## k-nearest neighbors algorithm (k-NN) is a non-parametric method used for classification and regression (wikipedia)

```{r, warning=FALSE}

library(HiDimDA)
data("AlonDS")
library(class)

set.seed(1234)
ind <- sample(2, nrow(AlonDS), replace=TRUE, prob=c(.39, 0.61))
apo.training <- AlonDS[ind==1 , 2:2001]
apo.test <- AlonDS[ind==2 , 2:2001]
apo.trainlabels <- AlonDS[ind==1, 1]
apo.testlabels <- AlonDS[ind==2, 1]
apo_pred <- knn(apo.training, apo.test, apo.trainlabels, k=1)
table(apo_pred, apo.testlabels)

```

## PCA dimension reduction

```{r}
alon.data <- (AlonDS[1:dim(AlonDS)[1], 2:dim(AlonDS)[2]])
alon.names <- read.csv("data/alon.names.csv", header=F)
Alon.cl <- c(0,1,0,1,0,1,0,1,0,1,0,1,0,1,0,1,0,1,0,1,0,1,0,1,0,0,0,0,0,0,0,0,0,0,0,0,0,0,1,0,0,1,1,0,0,0,0,1,0,1,1,0,0,1,1,0,0,0,0,1,0,1)
AlonFactor <- factor(Alon.cl,levels=0:1, labels= c("colonc","healthy"))
talon.data <- t(alon.data)
eigen(cor(talon.data))$values[1:5]
data <- talon.data; p <- ncol(data); n <- nrow(data) ; nboot<-1000
eigenvalues <- array(dim=c(nboot,p))
for (i in 1:nboot) {
    dat.star <- data[sample(1:n,replace=TRUE),]
    eigenvalues[i,] <- eigen(cor(dat.star))$values
}
for (j in 1:5) {
    print(quantile(eigenvalues[,j],c(0.025,0.975)))
}


```

## PCA results

```{r}
sum(eigen(cor(talon.data))$values[1:5])/62*100
#Thus, the first five components represent more than 87% of the variance in the data. The data can allow for a reduction in dimensions from 
#sixty two data points to five, while still capturing over 87% of the total variance

# to verify that the correlations between the patients are positive
# we run;
-eigen(cor(talon.data))$vec[,1:5]


```


## PCA results in five dimensions
```{r}
#From the previous slide we can match up the negative and positive signs with what we actually observed and express as;
Alon.cl
#By matching these we find that the 3rd and 4th dimensions matched the real data best. We can plot these with a biplot to show vectors
biplot(princomp(data,cor=TRUE),choices = c(3,4), pc.biplot=TRUE,cex=0.5,expand=0.8, main = "PCA Clustering of Alon et al. Data")
#There is a clear seperation from the left and right in the biplot

```

## PCA results; finding the top genes 
```{r}
#No we can prints the first ten gene names with respect to the third principal comonent to find out what the main genes of differentiating diseased and healthy state
pca <- princomp(talon.data, center = TRUE, cor=TRUE, scores=TRUE) 
o <- order(pca$scores[,3]) 
alon.names[o[1:10],6] 

pr.out =prcomp (alon.data , scale =TRUE)



```

## rpart finds important genes with recursive partioning and regression to make predictions 
```{r, warning=FALSE, message=FALSE}

library(rpart.plot)

row.names(talon.data)<- paste("gene", 1:2000, sep = "") 

alonRpart <- rpart(AlonFactor~., data=alon.data, method="class", cp=0.001) 
prp(alonRpart,  branch.lwd=4,  branch.col="blue",  extra=101, main="Recursive Partitioning and Regression Tree of Alon et al Data") 
alon.names[ 16, 6]
alon.names[ 24, 6]


```
## Random forests or random decision forests operate by constructing a multitude of decision trees at training time and outputting mean prediction (regression) of the individual trees. (wikipedia)
```{r, warning=FALSE, message=FALSE}
library(randomForest)
set.seed(1234)
alon.rf <- randomForest(grouping ~ ., data=AlonDS, importance=TRUE,
                      proximity=TRUE)

#We can predict the diseased state
 print(alon.rf)
 
```

##Random forest accuracy measures by best gene
```{r, warning=FALSE, message=FALSE}
 
## Random forest plot of gene accuracy 

 varImpPlot(alon.rf,
            n.var = 25,
            pch=19,
             main=NULL,
             col="blue",
             gcolor="blue",
             lcolor="black")
#Next we can find and print the genes that give the most accuracy
alon.names[249,6]
alon.names[1772,6]
alon.names[1671,6]

```

##Desmin gene prediction of diagnosis by expression level

```{r, warning=FALSE, message=FALSE}

colon.data <- read.table("data/AlonGene.txt")
alonLabels<-factor(Alon.cl, levels = 0:1, labels=c("colonc","healthy"));desmin <- grep("Human desmin gene, complete cds",alon.names[,6], ignore.case = TRUE)
alonPredictor <- factor(colon.data[desmin,]> 1700,levels=c("FALSE","TRUE"), labels=c("colonc","healthy"))
table(alonPredictor,alonLabels)

```

##We can also analyze a specific gene to measure its change of true positive or negative
```{r, warning=FALSE, message=FALSE}

library(ROCR)
alonLabels <- factor(Alon.cl,levels=0:1,labels= c("FALSE","TRUE"))
desmin <- grep("Human desmin gene, complete cds",alon.names[,6], ignore.case = TRUE)
  pred <- prediction(talon.data[desmin,], alonLabels)
  perf <- performance(pred, "tpr", "fpr" )
  plot(perf,main ="True Positive Rate of Human Desmin Gene",
         lwd=4,
        col="magenta")
performance(pred,"auc")

```

##K-means clustering is used after the top genes are found
```{r, warning=FALSE, message=FALSE}
RibosomalProtein <- grep("40S RIBOSOMAL PROTEIN S18",alon.names[,6], ignore.case = TRUE)
Desmin <- grep("Human desmin gene, complete cds",alon.names[,6], ignore.case = TRUE)
clusdata <- data.frame(talon.data[Desmin,],talon.data[RibosomalProtein,])
colnames(clusdata)<- c("Desmin", "40S Ribosomal Protein")
cl <- kmeans(clusdata, 2, 10)
plot(clusdata,    pch=as.numeric(AlonFactor) + 15,   col=as.numeric(AlonFactor) + 1,main = "K-means Center of Desmin and 40s Ribosomal Protein Clustering")
legend("topright",
           legend=c("Colon Cancer","Healthy"),
           pch=16:17,
           col=c(2,3))
 points(cl$centers, col=c(3,2), pch = 7, lwd=1)

```
##Here I use PAM clustering with the genes 40s Ribosomal Protein and Desmin
```{r, warning=FALSE, message=FALSE}
##Desmin gene prediction of diagnosis by expression level
library(cluster)
pamx <- pam(clusdata, 2)
plot(pamx,main="Pam Clustering by 40s Ribosomal Protein and Desmin")
legend("topright", legend=c("Colon Cancer","Healthy"), pch=16:17, col=c(2,3))
```

